Boosted Mirror Descent, Frank-Wolfe, and Herding

### 2013-09-10 icml/paper.tex
changed format to unix from dos

Consider the following loss function:
\begin{equation}
L(u) = h(u) + \max_{\theta \in \Theta} [\langle \theta, u \rangle - R(\theta)]
\end{equation}
We can think of $h$ as a \emph{primal regularizer} and $R$ as a \emph{dual regularizer}.

@@

First, let us generalize the setting 
to consider a \emph{two-argument} loss function
\begin{equation}
L(u,\theta) = h(u) + \langle \theta, u \rangle - R(\theta).
\end{equation}
We will assume throughout that $h$ and $R$ are both convex functions, 
and furthermore that $\arg\max_{u} h(u) + \theta^Tu$ can be efficiently 
computed for all values of $\theta$.

Mirror descent, together with boosting, yields an algorithm for finding 
``saddle points'' of $L$. It is the following algorithm (called Algorithm 1).
For notational convenience, for a sequence of weights $\alpha_1,\alpha_2,\ldots$ 
let $\hat{u}_t = \frac{\sum_{s=1}^t \alpha_su_s}{\sum_{s=1}^t \alpha_s}$ and let 
$\hat{\theta}_t = \frac{\sum_{s=1}^t \alpha_s\theta_s}{\sum_{s=1}^t \alpha_s}$.
\begin{enumerate}
\item $u_1 \in \arg\min_u h(u)$
\item $\theta_{t} \in \arg\max_{\theta \in \Theta} \langle \theta, u_t \rangle - R(\theta)$
\item $u_{t+1} \in \arg\min_{u} h(u) + \langle \hat{\theta}_t, u \rangle$
%\item $u_{t+1} \in \arg\min_{u} h(u) + \langle \frac{1}{t} \sum_{s \leq t} \theta_s, u \rangle$
\end{enumerate}
As long as $h$ is strongly convex, we obtain the 
bound (see Proposition~\ref{cor:method-1}):
\begin{equation}
\sup_{\theta \in \Theta} L(\hat{u}_T, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(1/T).
\end{equation}
In other words, $\hat{u}_T$ is close to being a global minimum of $L$.
However, it is often the case that $h$ is not strongly convex, whereas $R$ is strongly convex. In this case 
we may wish to use the following slightly different algorithm (called Algorithm 2):
\begin{enumerate}
\item $\theta_1 \in \arg\min_{\theta} R(\theta)$
\item $u_t \in \arg\min_{u} h(u) + \langle \theta_t, u \rangle$
%\item $\theta_{t+1} \in \arg\max_{\theta \in \Theta} \langle \theta, \frac{1}{t} \sum_{s \leq t} u_s \rangle - R(\theta)$
\item $\theta_{t+1} \in \arg\max_{\theta \in \Theta} \langle \theta, \hat{u}_t \rangle - R(\theta)$
\end{enumerate}
We then obtain the following bound 
when $R$ is strongly convex (see Proposition~\ref{cor:method-2}):
\[ \sup_{\theta \in \Theta} L(\hat{u}_T, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(1/T). \]
Algorithm 1 and Algorithm 2 are closely related; indeed, they are dual to each other 
(performing Algorithm 1 on $L(u,\theta)$ is the same as performing Algorithm 2 on 
$-L(\theta,u)$).

@@

Suppose that $R$ is $\alpha$-strongly convex and let $R = \sup_{u} |\langle u, u \rangle|$. 
\[ \sup_{\theta} L(\hat{u}, \theta) \leq \sup_{\theta} L(u^*, \theta) + \frac{4R^2\log(T+1)}{\alpha T}. \]

@@

\section{Application to Herding}
\label{sec:herding}
Suppose we are given a family of features $\phi$ together 
with a known value $\bar{\phi}$ for $\bE_{\mu}[\phi]$. We 
would like to construct a distribution that approximately 
matches this distribution, i.e. one for which 
$\bE_{\hat{\mu}}[\phi] \approx \bar{\phi}$. A natural way 
to do this is by using boosted mirror descent to minimize the 
maximum mean discrepancy relative to $\phi$; 
in other words, to minimize
\[ L(\mu) = \max\limits_{i=1}^d |\bE_{\mu}[\phi_i]-\bar{\phi}_i|. \]
If we wish to write this more similarly to the original problem 
formulation, we can write
\[ L(\mu) = \sup_{\|\theta\|_1 \leq 1} \theta^T(\bE_{\mu}[\phi]-\bar{\phi}). \]
One issue is that $\|\theta\|_1$ is not strongly convex, so we relax $\|\theta\|_1$ 
to $\|\theta\|_2$ and write our loss function in the Lagrangian form:
\[ L(\mu, \theta) = \theta^T\bE_{\mu}[\phi] - \theta^T\bar{\phi} - \frac{1}{2}\|\theta\|_2^2. \]
Intriguingly, Algorithm 1 and Algorithm 2 are identical to each other for this choice of $L$ 
(up to a change of index). In this case, $\mu_t = \delta_{x_t}$ for some $x_t$, and 
the updates are the same as the standard herding updates. Let $R = \sup_{x} \|\phi(x)\|_2$; 
then using the Algorithm 2 convergence bound gives us
\begin{align*}
\sup_{\theta} L\left(\frac{1}{T} \sum_{t=1}^T \delta_{x_t}, \theta\right) &\leq \inf_{u^*} \sup_{\theta} L(u^*, \theta) + \frac{2R^2\log(T+1)}{T} \\
 &= \sup_{\theta} \inf_{u^*} L(u^*, \theta) + \frac{2R^2\log(T+1)}{T} \\
 &= \frac{2R^2\log(T+1)}{T}.
\end{align*}
This is a slightly weaker version of the typical herding bound on MMD.

@@

Let us now consider the entropic regularizer $h(\mu) = \bE_{\mu}[\log \mu(x)]$ with $R(\theta) = \bE_{\mu}[\theta^T\bar{\phi}] + \left\{ \begin{array}{llc} \infty & : & \|\theta\|_1 > 1 \\ 0 & : & \|\theta\|_1 \leq 1 \end{array} \right.$.
If we use Method 1, we then end up with 
\[ \mu_i(x) \propto \exp(\frac{1}{i}(\sum_{j<i} \theta_j)^T\phi(x)), \]
so $h(\mu) + \bE_{\mu}[\theta^T\phi(x)] - R(\theta)$ is no longer maximized at the 
boundary. We can instead consider the following sequence of distributions:
\begin{enumerate}
\item $\theta_i \in \argmax_{\|\theta\|_1 \leq 1} \theta^T (\bE_{\mu_i}[\phi(x)]-\bar{\phi})$
\item $\mu_{i+1} \propto \exp(\frac{1}{i}(\sum_{j \leq i} \theta_j)^T\phi(x))$
\end{enumerate}
The entropic regularizer is not strongly convex but a more general analysis involving 
Bregman divergence can nevertheless yield a convergence bound in this case. We thus 
get something that looks like herding but actually gives us an exponential family 
type distribution.

>>>>

\NA{JHH: I'm thinking that from the beginning we should let $u$ be defined over a different space than $\theta$ and assume a linear map $u \mapsto \eta(u) \in \Theta$. Then write $\langle \theta, \eta(u) \rangle$ so the setup will align with the herding case, where $\eta(u) = \bE_{u}[\phi(x)]$. }

Consider the following loss function:
\begin{equation}
L(u) = h(u) + \max_{\theta \in \Theta} \left\{\langle \theta, u \rangle - R(\theta) \right\}.
\end{equation}
We can think of $h$ as a \emph{primal regularizer} and $R$ as a \emph{dual regularizer}.

@@

\NA{JHH: Should we comment on the symmetry introduced by this formulation?}
First, let us generalize the setting 
to consider a \emph{two-argument} loss function
\begin{equation}
L(u,\theta) = h(u) + \langle \theta, u \rangle - R(\theta).
\end{equation}
We will assume throughout that $h$ and $R$ are both convex functions, 
and furthermore that $\arg\max_{u} h(u) + \theta^Tu$ can be efficiently 
computed for all values of $\theta$.

Mirror descent, together with boosting, yields an algorithm for finding 
``saddle points'' of $L$. It is the following algorithm (called Algorithm 1).
For notational convenience, for a sequence of weights $\alpha_1,\alpha_2,\ldots$ 
let $\hat{u}_t = \frac{\sum_{s=1}^t \alpha_su_s}{\sum_{s=1}^t \alpha_s}$ and let 
$\hat{\theta}_t = \frac{\sum_{s=1}^t \alpha_s\theta_s}{\sum_{s=1}^t \alpha_s}$.
\begin{enumerate}
\item $u_1 \in \arg\min_u h(u)$
\item $\theta_{t} \in \arg\max_{\theta \in \Theta} \langle \theta, u_t \rangle - R(\theta)$
\item $u_{t+1} \in \arg\min_{u} h(u) + \langle \hat{\theta}_t, u \rangle$
%\item $u_{t+1} \in \arg\min_{u} h(u) + \langle \frac{1}{t} \sum_{s \leq t} \theta_s, u \rangle$
\end{enumerate}
As long as $h$ is strongly convex, we obtain the 
bound (see Proposition~\ref{cor:method-1}):
\begin{equation}
\sup_{\theta \in \Theta} L(\hat{u}_T, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(1/T).
\end{equation}
In other words, $\hat{u}_T$ is close to being a global minimum of $L$.
However, it is often the case that $h$ is not strongly convex, whereas $R$ is strongly convex. In this case 
we may wish to use the following slightly different algorithm (called Algorithm 2):
\begin{enumerate}
\item $\theta_1 \in \arg\min_{\theta} R(\theta)$
\item $u_t \in \arg\min_{u} h(u) + \langle \theta_t, u \rangle$
%\item $\theta_{t+1} \in \arg\max_{\theta \in \Theta} \langle \theta, \frac{1}{t} \sum_{s \leq t} u_s \rangle - R(\theta)$
\item $\theta_{t+1} \in \arg\max_{\theta \in \Theta} \langle \theta, \hat{u}_t \rangle - R(\theta)$
\end{enumerate}
We then obtain the following bound 
when $R$ is strongly convex (see Proposition~\ref{cor:method-2}):
\[ \sup_{\theta \in \Theta} L(\hat{u}_T, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(1/T). \]
Algorithm 1 and Algorithm 2 are closely related; indeed, they are dual to each other 
(performing Algorithm 1 on $L(u,\theta)$ is the same as performing Algorithm 2 on 
$-L(\theta,u)$).
\NA{JHH: It would be more elegant to argue either here or below that the bounds on the second algorithm follow directly from duality}

@@

Suppose that $R$ is $\gamma$-strongly convex and let $r = \sup_{u} |\langle u, u \rangle|$. Then

@@

\section{Application to Herding}
\label{sec:herding}

\NA{JHH: Need to specify the spaces we are working with here.}
Suppose we are given a family of features $\phi$ together 
with a known value $\bar{\phi}$ for $\bE_{\mu}[\phi]$. We 
would like to construct a distribution that approximately 
matches this distribution, i.e.~one for which 
$\bE_{\hat{\mu}}[\phi] \approx \bar{\phi}$. A natural way 
to do this is by using boosted mirror descent to minimize the 
maximum mean discrepancy relative to $\phi$; 
in other words, to minimize
\[ L(\mu) = \max_{i \in \{1,\dots,d\}} |\bE_{\mu}[\phi_i]-\bar{\phi}_i|. \]
If we wish to write this more similarly to the original problem 
formulation, we can write
\[ L(\mu) = \sup_{\|\theta\|_1 \leq 1} \theta^T(\bE_{\mu}[\phi]-\bar{\phi}). \]
One issue is that $\|\theta\|_1$ is not strongly convex, so we relax $\|\theta\|_1$ 
to $\|\theta\|_2$ and write our loss function in the Lagrangian form:
\[ L(\mu, \theta) = \theta^T\bE_{\mu}[\phi] - \theta^T\bar{\phi} - \frac{1}{2}\|\theta\|_2^2. \]
Intriguingly, Algorithm 1 and Algorithm 2 are identical to each other for this choice of $L$ 
(up to a change of index). In this case, $\mu_t = \delta_{x_t}$ for some $x_t$, and 
the updates are the same as the standard herding updates. Let $r = \sup_{x} \|\phi(x)\|_2$; 
then using the Algorithm 2 convergence bound gives us
\begin{align*}
\sup_{\theta} L\left(\frac{1}{T} \sum_{t=1}^T \delta_{x_t}, \theta\right) &\leq \inf_{u^*} \sup_{\theta} L(u^*, \theta) + \frac{2R^2\log(T+1)}{T} \\
 &= \sup_{\theta} \inf_{u^*} L(u^*, \theta) + \frac{2R^2\log(T+1)}{T} \\
 &= \frac{2R^2\log(T+1)}{T}.
\end{align*}
This is a slightly weaker version of the typical herding bound on MMD.

@@

Let us now consider the entropic regularizer $h(u) = \tau\bE_{u}[\log u(x)]$ with 
\[ R(\theta) = \bE_{u}[\theta^T\bar{\phi}] + \left\{ \begin{array}{llc} \infty & : & \|\theta\|_1 > 1 \\ 0 & : & \|\theta\|_1 \leq 1 \end{array} \right.\]
If we use Method 1, we then end up with 
\[ u_{t}(x) \propto \exp\left\{-\frac{1}{\alpha}\overline{\theta}_{t}^T\phi(x)\right\}, \]
where $\overline{\theta}_{t} = \frac{1}{t} \sum_{s \le t} \theta_{s}$. Hence, $h(u) + \bE_{\mu}[\theta^T\phi(x)] - R(\theta)$ is no longer maximized at the boundary (when $\tau > 0$). We can instead consider the following sequence of distributions:
\begin{enumerate}
\item $\theta_t \in \argmax_{\|\theta\|_1 \leq 1} \theta^T (\bE_{u_t}[\phi(x)]-\bar{\phi})$
\item $u_{t+1}(x) = Z(\overline{\theta}_{t}, \tau)^{-1} \exp\left\{-\frac{1}{\tau}\overline{\theta}_{t}^T\phi(x)\right\}$
\end{enumerate}
Here $Z(\overline{\theta}_{t}, \tau) = \int\exp\left\{-\frac{1}{\tau}\overline{\theta}_{t}^T\phi(x)\right\} \dee x$.

The entropic regularizer is not strongly convex but a more general analysis involving 
Bregman divergence can nevertheless yield a convergence bound in this case. We thus 
get something that looks like herding but actually gives us an exponential family 
type distribution. In particular, we have a Gibbs distribution with temperature $\tau$, and 
in the limit as the temperature goes to zero, we recover herding, since 
\[ \lim_{\tau \to 0} u_{t+1}(x) = \delta_{x^{*}_{t+1}}, \]
where $x^{*}_{t+1} = \argmin_{x} \overline{\theta}_{t}^T\phi(x)$. 
This fact is reminiscent of Welling's original derivation of herding, in which 
he took the zero temperature limit of the gradient descent algorithm for the optimization problem 
\[ \argmin_{u} h(u) \quad \text{s.t.}~\bE_{u}[\phi(x)] - \bar\phi = 0. \]
This optimization is quite similar to the entropic optimization considered here, which 
we can write as 
\begin{align*}
\argmin_{u}& \argmax_{|\theta\|_1 \leq 1} h(u) + \theta^{T}(\bE_{u}[\phi(x)] - \bar \phi)  \\
&= \argmin_{u} h(u) + \|\bE_{u}[\phi(x)] - \bar\phi\|_{2}. 
\end{align*}
In the entropic formulation, the expectation constraint $\bE_{u}[\phi(x)] - \bar\phi = 0$ becomes
soft, though the two optimizations are equivalent in the zero-temperature limit.

We can view the entropic regularization
with non-zero temperature as an alternative approximation to herding. If at each step we
retain a sample $x_{t} \sim u_{t}$, then the sequence $(x_{t})_{t}$ converges to the 
herding sample path as $\tau \to 0$. So $(x_{t})_{t}$ with non-zero temperature approximate
herding samples. \NA{JHH: I'm not sure how useful this all is, other than to make contact with 
the original paper, as I think the Gibbs distribution was exactly what Max was trying to 
avoid by taking the zero-temperature limit.}
